A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2020-08-19 new-paper.Rmd
adding some of the missing references

Meaningful progress on computational reproducibility, the ability of an independent researcher to recreate figures, tables and key findings, became more of a practical reality in the late 2000s as new tools and services came into existence. Scientists started sharing more code and software, partly due to the rapid increase in training opportunities made possible by organizations like The Carpentries, combined with the growing adoption of Github by scientists [@Ram2013; @Zagalsky2015]. The bigger driver for this is more likely driven by the growing popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has been catalyzed by a Cambrianesque explosion of open source software tools. Programming languages such as Python, R and Julia have helped scientists implement and share new methods to work with data [@RCore; @python; @julia]. Each of these languages enjoys support of thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. Modern scientists now have access to many interactive computational frameworks (e.g. Jupyter, RMarkdown, Stencila, Observable) that can interact with kernels from dozens of programming languages. Beyond notebooks to run the code, a reader will also require access to the same software dependencies. In a study of applied computational research, [@Collberg2014] and colleagues were only able to install and run less than 20% of the software described in publications. In addition to locating software, one still has to install the right versions. Although this problem is far from solved, technologies like Docker have alleviated some of these concerns. Researchers can share Dockerfiles to recreate computational environments [@Boettiger2015]. Services like Binder can then leverage Docker and make possible to share the computational environments in which these notebooks were run. The barriers to verification are now low enough that trivial reproducibility is sometimes a click away, free from the misery of software installation [@Jupyter2018].

@@

Tools to prepare datasets and metadata (*the infrastructure*) and best practices (guidelines) for reuse are sparse [@Arslan2019; @Roche2015]. Unlike code, datasets are far more diverse in terms of heterogeneity, size, and formats. This makes them particularly challenging to standardize or easily "install" where the code is running. There are some efforts working towards improving this process, such as gogetdata , the 'dat' project, and a private company, Quilt [@goget; @dat; @Quilt].

@@

To make data more reusable, we need to make it easier to generate metadata, standardise good practices within and across communities, and make reusable data sharing a recognized, normalised, part of research. Here we present a set of best practices to enable this change. These include documentation (\@ref(readme)), licensing (\@ref(licenses)), citation (\@ref(cite)), code (\@ref(pkgs)), raw and analysis ready data (\@ref(rawdata)). We also discuss considerations for sharing of data of various sizes \@ref(size).

@@

A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. We recommend using a license from Creative Commons [@creativecommons], specifically, we recommend CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. We recommend against non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

@@

Metadata should provided in plain text files following a standard format. Such formats include the Ecological Metadata Language [@EML-about] (EML), which uses XML (The eXtensible Markup Language), and Table Schema, which uses JSON-LD (Java Script Object Notation - Linked Data) [@ref](fig:json-ld). One implementation of this is Frictionless Data [@Fowler2017], which provides a language agnostic way to document data - to learn more see [@frictionless-data-guide].

EML is widely used in Ecology (for example, <https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-cwt.3091.13>). JSON-LD provides advantages over EML, by having fields to facilitate data discovery with google dataset search [@EML-data]. We recommend JSON-LD over EML, as it is more human readable, shorter, and quicker and easier to read and write. For more information on tooling to assist generating metadata, see <https://knb.ecoinformatics.org/tools/morpho> and [@Arslan2019].

@@

Citation is important for provenance, and credit. Citing data clarifies the data version used (provenance), and provides authors a citation (credit). To ensure these are provided, data needs to have an accession number, such as a Digital Object Identifier (DOI). A DOI is a permanent, unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. If a DOI or accession number is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named CITATION should be placed in the directory at the top level. It should contain a DOI, and could be in `.bibtex` format - [\@ref(fig:dir-layout). Avenues that provide DOIs are discussed in \@ref(avenues-to-share-data)].

@@

As discussed in \@ref(cite), we recommend data are archived using a platform that generates an accession number such as a DOI. For sharing data collected for research, the most relevant homes are the research data repositories **zenodo**, **dryad**, or **Open Science Framework** (OSF). In addition to minting DOIs, they also provide citation templates that can be used in \@ref(cite).

@@

Data can be distributed alongside code using a format such as a package in R, Python, and Julia [@nycflights; @penguins; @python-data; @r-data-julia]. A package or module structure provides an advantage as it can be "installed" and loaded just like code. Despite the advantages of sharing data as a software package, there are a few major disadvantages. Firstly, packaging data inside a software package makes the data availability language centric. Users who work in other languages are unlikely to download and export data out of a package. Secondly, data size can be a limitation depending on the repository (CRAN, the R repository, for example, limits package sizes to 5mb although there are workarounds [@Brooke_Anderson2017]). Other fields, such as bioinformatics have a process of sharing data through repositories like Bioconductor, with larger data size. Thirdly, there is the cost of authors and data curators/custodians understanding intricacies of package development in order to share data.

>>>>

Meaningful progress on computational reproducibility, the ability of an independent researcher to recreate figures, tables and key findings, became more of a practical reality in the late 2000s as new tools and services came into existence. Scientists started sharing more code and software, partly due to the rapid increase in training opportunities made possible by organizations like The Carpentries, combined with the growing adoption of Github by scientists [@Ram2013; @Zagalsky2015]. The bigger driver for this is more likely driven by the growing popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has been catalyzed by a Cambrianesque explosion of open source software tools. Programming languages such as Python, R and Julia have helped scientists implement and share new methods to work with data [@rcore; @python; @julia]. Each of these languages enjoys support of thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. Modern scientists now have access to many interactive computational frameworks (e.g. Jupyter, RMarkdown, Stencila, Observable) that can interact with kernels from dozens of programming languages. Beyond notebooks to run the code, a reader will also require access to the same software dependencies. In a study of applied computational research, [@Collberg2014] and colleagues were only able to install and run less than 20% of the software described in publications. In addition to locating software, one still has to install the right versions. Although this problem is far from solved, technologies like Docker have alleviated some of these concerns. Researchers can share Dockerfiles to recreate computational environments [@Boettiger2015]. Services like Binder can then leverage Docker and make possible to share the computational environments in which these notebooks were run. The barriers to verification are now low enough that trivial reproducibility is sometimes a click away, free from the misery of software installation [@Jupyter2018].

@@

Tools to prepare datasets and metadata (*the infrastructure*) and best practices (guidelines) for reuse are sparse [@Arslan2019; @Roche2015]. Unlike code, datasets are far more diverse in terms of heterogeneity, size, and formats. This makes them particularly challenging to standardize or easily "install" where the code is running. There are some efforts working towards improving this process, such as gogetdata , the 'dat' project, and a private company, Quilt [@goget; @dat; @quilt].

@@

To make data more reusable, we need to make it easier to generate metadata, standardise good practices within and across communities, and make reusable data sharing a recognized, normalised, part of research. Here we present a set of best practices to enable this change. These include documentation (\@ref(documentation-readmes)), licensing (\@ref(licences)), citation (\@ref(citation-how-you-want-your-data-to-be-cited)), code (\@ref(data-as-part-of-open-source-packages)), raw and analysis ready data (\@ref(raw-data-vs-analysis-read-data)). We also discuss considerations for sharing of data of various sizes \@ref(considerations-for-data-size).

@@

A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. We recommend using a license from Creative Commons (<https://creativecommons.org/>), specifically, we recommend CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. We recommend against non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

@@

Metadata should provided in plain text files following a standard format. Such formats include the Ecological Metadata Language [@EML-about] (EML), which uses XML (The eXtensible Markup Language), and Table Schema, which uses JSON-LD (Java Script Object Notation - Linked Data) \@ref(fig:json-ld). One implementation of this is Frictionless Data [@Fowler2017], which provides a language agnostic way to document data - to learn more see <https://frictionlessdata.io/guide/>.

EML is widely used in Ecology (for example [@Maerz2014]). JSON-LD provides advantages over EML, by having fields to facilitate data discovery with google dataset search [@EML-data]. We recommend JSON-LD over EML, as it is more human readable, shorter, and quicker and easier to read and write. For more information on tooling to assist generating metadata, see <https://knb.ecoinformatics.org/tools/morpho> and [@Arslan2019].

@@

Citation is important for provenance, and credit. Citing data clarifies the data version used (provenance), and provides authors a citation (credit). To ensure these are provided, data needs to have an accession number, such as a Digital Object Identifier (DOI). A DOI is a permanent, unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. If a DOI or accession number is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named CITATION should be placed in the directory at the top level. It should contain a DOI, and could be in `.bibtex` format - \@ref(fig:dir-layout). Avenues that provide DOIs are discussed in \@ref(avenues-to-share-data)].

@@

As discussed in \@ref(citation-how-you-want-your-data-to-be-cited), we recommend data are archived using a platform that generates an accession number such as a DOI. For sharing data collected for research, the most relevant homes are the research data repositories **zenodo**, **dryad**, or **Open Science Framework** (OSF). In addition to minting DOIs, they also provide citation templates that can be used in \@ref(citation-how-you-want-your-data-to-be-cited).

@@

Data can be distributed alongside code using a format such as a package in R, Python, and Julia [@nycflights; @horst-penguins; @data-python-pkg; @r-data-julia]. A package or module structure provides an advantage as it can be "installed" and loaded just like code. Despite the advantages of sharing data as a software package, there are a few major disadvantages. Firstly, packaging data inside a software package makes the data availability language centric. Users who work in other languages are unlikely to download and export data out of a package. Secondly, data size can be a limitation depending on the repository (CRAN, the R repository, for example, limits package sizes to 5mb although there are workarounds [@Brooke_Anderson2017]). Other fields, such as bioinformatics have a process of sharing data through repositories like Bioconductor, with larger data size. Thirdly, there is the cost of authors and data curators/custodians understanding intricacies of package development in order to share data.
